Effectful Semantics for Substructural SSA

### 2023-11-15 pldi24/paper.tex
Tap

We're now ready to give a semantics for an \isotopessa dialect with a set of
types \(\mc{T}\) and instructions \(\mc{I}\). We'll need:
\begin{itemize}
  \item A \textit{substructural effectful category} \(\mc{C} = \mc{C}_1 \to
  \mc{C}_0\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot},
  equipped with
  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects
  in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies \ms{base}(X) \in
  \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies \ms{base}(X) \in
  \ms{Rel}(\mc{C})\)
  \item A function \(\ms{inst}(f)\) from instructions \(f \in \mc{I}_p^q(A, B)\)
  to morphisms in \(\mc{C}_p^q(\dnt{A}, \dnt{B})\)
\end{itemize}
We begin by giving a semantics for types and contexts in terms of objects in
\(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}: we map tensor products to
tensor products, the unit type to the monoidal unit, and booleans to \(\mb{1} +
\mb{1}\). Contexts are simply interpreted as the tensor products of the variable
types which appear in them; in particular, the quantity annotations have
\textit{no} effect on a context's semantics. We then give a semantics for
structural judgements in Figure \ref{fig:ssa-structural-semantics}. In
particular, we begin by interpreting the judgement that a type is affine,
\(\tlin{A}{\taff}\), as the drop morphism \(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A},
I)\) allowing us to discard the type by sending it to the monoidal unit; type
being relevant, \(\tlin{A}{\trel}\) is similarly interpreted as the split
morphism \(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, \dnt{A} \otimes \dnt{A})\). We
can then interpret a weakening \(\csplits{\Gamma}{\Delta}{\Xi}\) as a pure
morphism \(\mc{C}^\tint_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})\) which
partitions the variables as appropriate, splitting and dropping as necessary;
weakenings are then just interpreted as the postcomposition of the associated
split morphism and the unitor.

>>>>

We're now ready to give a semantics for an \isotopessa dialect with a set of
types \(\mc{T}\) and instructions \(\mc{I}\). We'll need:
\begin{itemize}
  \item A \textit{substructural effectful category} \(\mc{C} = \mc{C}_1 \to
  \mc{C}_0\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot},
  equipped with
  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects
  in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies \ms{base}(X) \in
  \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies \ms{base}(X) \in
  \ms{Rel}(\mc{C})\)
  \item A function \(\ms{inst}(f)\) from instructions \(f \in \mc{I}_p^q(A, B)\)
  to morphisms in \(\mc{C}_p^q(\dnt{A}, \dnt{B})\)
\end{itemize}
We begin by giving a semantics for types and contexts in terms of objects in
\(|\mc{C}|\) in Figure \ref{fig:ssa-type-semantics}: we map tensor products to
tensor products, the unit type to the monoidal unit, and booleans to \(\mb{1} +
\mb{1}\). Contexts are simply interpreted as the tensor products of the variable
types which appear in them; in particular, the quantity annotations have
\textit{no} effect on a context's semantics. We then give a semantics for
structural judgements in Figure \ref{fig:ssa-structural-semantics}. In
particular, we begin by interpreting the judgements that a type is affine (i.e.
\(\tlin{A}{\taff}\)) or relevant (i.e. \(\tlin{A}{\trel}\)) as the drop morphism
\(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, I)\) and split morphism
\(\ms{drop}(\dnt{A}): \mc{C}(\dnt{A}, \dnt{A} \otimes \dnt{A})\) respectively.
We can then interpret a weakening \(\csplits{\Gamma}{\Delta}{\Xi}\) as a pure
morphism \(\mc{C}^\tint_1(\dnt{\Gamma}, \dnt{\Delta} \otimes \dnt{\Xi})\) which
partitions the variables as appropriate, splitting and dropping as necessary;
weakenings are then just interpreted as the postcomposition of the associated
split morphism and the unitor.
